Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-11-17 draft_lm1.tex
added table for attack ratios; updated the figure for the estimation; started writing the results;

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.

@@

The time series used for all the analyses consist of officially notified dengue 
cases agregated weekly.
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$. The authors are also grateful to Claudia T. 
Codeco for helpful discussions about the manuscript.

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{./plots/concat_SI.png}
\end{center}
\caption{
{\bf Bold the first sentence.}  Rest of figure caption.  
}
\label{Fig:S0}
\end{figure}

@@

>>>>

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
In this study we used one chain per parameter, i.e, 3 chains for each run.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubin's R convergence diagnostic~\cite{brooks1998}.

@@

The time series used for all the analyses consists of officially notified 
dengue cases aggregated weekly. 
Figure XX  shows the $R_t$ series, according to 
(\ref{eq:Rtestimate}). 
It can be seen that the inter-epidemic periods are characterized by $R_t$ being 
indistinguishable from 1.

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data.

Table \ref{tab:AR} contains the attack ratios calculated for each 
epidemic/outbreak.
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$. The authors are also grateful to Claudia T. 
Codeco for helpful discussions about the manuscript.

\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\begin{center}
\includegraphics[width=.9\textwidth]{./concat_SI.png}
\end{center}
\caption{
{\bf Susceptibles and Infectious posterior curves.}  The curves were estimated 
only for the periods where $R_t> 1$.  The susceptible curves in the top panel 
reflect the prevalence of fraction of susceptibles to circulating strain(s) for 
each epidemic/outbreak. In the lower panel see the posterior distribution 
of infectious curves, represented by its Median and 95\% credible interval. 
Scatter plot shows the observed cases.
}
\label{Fig:S0}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Median Attack ration calculated according to (\ref{eq:AR2})}}
\begin{center}
\begin{tabular}{l|r}
\hline
Year & Median Attack Ratio \\
\hline
1996 & \textbf{38.64\%} (17.18\%-54.19\%)\\
1997 & \textbf{87.31\%} (73.83\%-87.31\%)\\
1998 & \textbf{49.86\%} (49.19\%-49.86\%)\\
1999 & \textbf{10.76\%} (3.72\%-20.24\%)\\
2000 & \textbf{25.49\%} (24.47\%-26.98\%)\\
2001 & \textbf{48.12\%} (46.57\%-49.15\%)\\
2005 & \textbf{14.74\%} (10.23\%-21.38\%)\\
2006 & \textbf{11.22\%} (8.00\%-14.41\%)\\
2007 & \textbf{15.08\%} (14.76\%-15.38\%)\\
2008 & \textbf{13.73\%} (3.10\%-30.84\%)\\
2010 & \textbf{18.30\%} (17.34\%-19.33\%)\\
2011 & \textbf{8.59\%} (8.20\%-9.36\%)\\
2012 & \textbf{13.92\%} (12.71\%-14.92\%)\\
\hline
\end{tabular}
\begin{flushleft}Medians are presented with 95\% credible intervals.
\end{flushleft}

\end{center}

\label{tab:AR}
\end{table}
